---
title: "analysis"
author: "Eliot Min"
date: "7/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot)
library(stringr)
library(tidyverse)
```

```{r reading in data}

data <- read_csv("MLB statcast data/stats.csv")

```


# Question 1: The ability to "hit to all fields" is a very highly touted skill,
# often the sign of a mature player. Should this be taken into account to
# explain the discrepancy between expected batting average and actual batting
# average?

# Will start out by doing an analysis of the 2017 data:

```{r}

data_2017 <- data %>%
  select(last_name, first_name, player_id, year, batting_avg, xba, pull_percent, straightaway_percent, opposite_percent) %>%
  filter(year == 2017) %>%
  mutate(difference = batting_avg-xba)

  ggplot(data_2017, aes(x = pull_percent, y = difference)) + geom_point() + 
  labs(x = "Pull Percent", 
       y = "Difference between batting average and expected batting average") +
    geom_smooth(method = 'lm') +
  theme_classic()

```

#Now the 2018 data:


```{r}

data_2018 <- data %>%
  select(last_name, first_name, player_id, year, batting_avg, xba, pull_percent, straightaway_percent, opposite_percent) %>%
  filter(year == 2018) %>%
  mutate(difference = batting_avg-xba)

  ggplot(data_2018, aes(x = pull_percent, y = difference)) + geom_point() + 
  labs(x = "Pull Percent", 
       y = "Difference between batting average and expected batting average") +
    geom_smooth(method = 'lm') +
  theme_classic()

```

#2019 data:


```{r}

data_2019 <- data %>%
  select(last_name, first_name, player_id, year, batting_avg, xba, pull_percent, straightaway_percent, opposite_percent) %>%
  filter(year == 2019) %>%
  mutate(difference = batting_avg-xba)

  ggplot(data_2019, aes(x = pull_percent, y = difference)) + geom_point() + 
  labs(x = "Pull Percent", 
       y = "Difference between batting average and expected batting average") +
    geom_smooth(method = 'lm') +
  theme_classic()

```

#2020 and 2021 data to be combined due to the shorter season thus far:


```{r}

data_2020_2021 <- data %>%
  select(last_name, first_name, player_id, year, batting_avg, xba, pull_percent, straightaway_percent, opposite_percent) %>%
  filter(year %in% c(2020, 2021)) %>%
  mutate(difference = batting_avg-xba)

  ggplot(data_2020_2021, aes(x = pull_percent, y = difference)) + geom_point() + 
  labs(x = "Pull Percent", 
       y = "Difference between batting average and expected batting average") +
    geom_smooth(method = 'lm') +
  theme_classic()

```

#Fascinating... we see a negative correlation. Insofar as we want xBA to "indicate" whether a batter is getting lucky or unlucky, one might make the case that 
