---
title: "Should MLB weigh pull tendency when calculating xBA?"
author: "Eliot Min"
date: "7/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(ggplot2)
library(stringr)
library(tidyverse)
library(blogdown)

```


```{r reading in data, warning = FALSE, message = FALSE}

#Data is taken from baseball savant — thanks so much for making it available!

batting_data <- read_csv("MLB statcast data/stats.csv")

```

Expected batting average only takes into account the angle at which a batter
hits a ball, the velocity at which the ball is hit and, in certain cases, the
speed at which a batter is capable of running. However, it does not take into
account the *direction* the ball is hit in — that is, it does not take into
account how often a right-handed batter hits the ball to left field, his
pull-side, instead of to center or right field.

We'll do more in-depth analyses of each individual season, but for now, let's
look at the data altogether:

```{r, message=FALSE}

full_data <- batting_data %>%
  mutate(difference = batting_avg-xba)

  ggplot(full_data, aes(x = pull_percent, y = difference)) + 
    geom_point() + 
  labs(x = "Pull Percent", 
       y = "Batting Average - Expected Batting Average",
       title = "Difference between BA and xBA for hitters against pull percent \n 2017-2021") +
    geom_smooth(method = 'lm') +
    annotate("text", x = 30, y = .08, label = "r = .20") + 
  theme_classic()
  
  #model to gather significance values and correlation value
  
  full_data_model <- lm(difference ~ pull_percent, data = full_data)
  full_data_cor <- cor(full_data$difference, 
                       full_data$pull_percent, method = "pearson")
  
```

There is a statistically significant negative correlation between BA-xBA and
pull percentage (p<.001) — this would seem to suggest that the more you pull the
ball (and, resultingly, the less you use the whole field) the more likely you
are to underperform your xBA.

Let's see if this trend holds for individual years of xBA vs BA data. I will
start out by doing an analysis of the 2017 data:

```{r, message = FALSE}

ba_2017 <- batting_data %>%
  filter(year == 2017) %>%
  mutate(difference = batting_avg-xba)

  ggplot(ba_2017, aes(x = pull_percent, y = difference)) + 
    geom_point() + 
  labs(x = "Pull Percent", 
       y = "Batting Average - Expected Batting Average",
       title = "Difference between BA and xBA for hitters in 2017 as a function of pull %") +
    geom_smooth(method = 'lm') +
  theme_classic()

  cor_2017 <- cor(ba_2017$difference, ba_2017$pull_percent, method = "pearson")
  
  model_2017 <- lm(difference~pull_percent, data = ba_2017)

```

2018 data:

```{r,  message=FALSE}

data_2018 <- batting_data %>%
  select(last_name, first_name, player_id, year, batting_avg, xba, pull_percent, straightaway_percent, opposite_percent) %>%
  filter(year == 2018) %>%
  mutate(difference = batting_avg-xba)

  ggplot(data_2018, aes(x = pull_percent, y = difference)) + geom_point() + 
  labs(x = "Pull Percent", 
       y = "Difference between batting average and expected batting average") +
    geom_smooth(method = 'lm') +
  theme_classic()

  model_2018 <- lm(difference~pull_percent, data = data_2018)
```

2019 data:

```{r, message = FALSE}

data_2019 <- batting_data %>%
  select(last_name, first_name, player_id, year, batting_avg, xba, pull_percent, straightaway_percent, opposite_percent) %>%
  filter(year == 2019) %>%
  mutate(difference = batting_avg-xba)

  ggplot(data_2019, aes(x = pull_percent, y = difference)) + geom_point() + 
  labs(x = "Pull Percent", 
       y = "Difference between batting average and expected batting average") +
    geom_smooth(method = 'lm') +
  theme_classic()
  
  model_2019 <- lm(difference~pull_percent, data = data_2019)

```

2020 and 2021 data to be combined due to the shorter season in 2020 and the
current games played in 2021, which roughly equates to a full 162 games:

```{r, message=FALSE}

data_2020_2021 <- batting_data %>%
  select(last_name, first_name, player_id, year, batting_avg, xba, pull_percent, straightaway_percent, opposite_percent) %>%
  filter(year %in% c(2020, 2021)) %>%
  mutate(difference = batting_avg-xba)

  ggplot(data_2020_2021, aes(x = pull_percent, y = difference)) + geom_point() + 
  labs(x = "Pull Percent", 
       y = "Difference between batting average and expected batting average") +
    geom_smooth(method = 'lm') +
  theme_classic()
  
  
  model <- lm(difference~pull_percent, data = data_2020_2021)

```

We see a negative correlation for each year, although the correlation for the
2018 data is insignificant.

Since we want xBA to "indicate" whether a batter is getting lucky or unlucky,
one might make the case that the degree to which one is able to hit the ball to
the opposite field/straightaway is currently lacking in the xBA calculation. xBA
should probably be scaled upwards for hitters who pull the ball less often,
while xBA should be scaled down for hitters who are pull heavy.
